---
title: "Lab 8"
author: "Rachel Ramos"
format: html
execute: 
  echo: true
---



The goal of this lab is learn more about exploring missing data and writing
modular code.

```{r}
#| label: setup

library(tidyverse)
library(dplyr)
library(ggplot2)

trout_data <- read_csv("BlackfootFish.csv")
```

## The Data

This lab's data concerns mark-recapture data on four species of trout from the
Blackfoot River outside of Helena, Montana. These four species are
**rainbow trout (RBT)**, **westslope cutthroat trout (WCT)**, **bull trout**,
and **brown trout**.

Mark-recapture is a common method used by ecologists to estimate a population's
size when it is impossible to conduct a census (count every animal). This method
works by *tagging* animals with a tracking device so that scientists can track
their movement and presence.

## Data Exploration

The measurements of each captured fish were taken by a biologist on a raft in
the river. The lack of a laboratory setting opens the door to the possibility of
measurement errors.

**1a. Let's look for missing values in the dataset. Output ONE table that answers BOTH of the following questions:**

+ **How many observations have missing values?** 1796
+ **What variable(s) have missing values present?** weight

::: callout-tip
# You should use `across()`!
:::

```{r}
#| label: find-missing-values
# Count how many rows have *any* missing values
num_rows_with_na <- trout_data %>%
  filter(if_any(everything(), is.na)) %>%
  nrow()

# Get the number of missing values for each column
missing_by_column <- trout_data %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "missing_count") %>%
  filter(missing_count > 0)

# Add a row at the top to show total rows with any missing values
missing_summary <- bind_rows(
  tibble(variable = "Rows with any missing", missing_count = num_rows_with_na),
  missing_by_column
)

missing_summary
```



**1b. Using `map_int()`, produce a nicely formatted table of the number of missing values for each variable in the `fish` data that displays the same information as 1a** 



```{r}
#| label: map-missing-values-of-fish
missing_summary_map <- map_int(trout_data, ~sum(is.na(.))) %>% #count NAs for each column
  enframe(name = "variable", value = "missing_count") %>%
  filter(missing_count > 0)

missing_summary_map

```


**2. Create ONE thoughtful visualization that explores the frequency of missing values across the different years, sections, and trips.**

```{r }
#| label: visual-of-missing-values-over-time

# Subset data with missing weight
missing_weight <- trout_data %>%
  filter(is.na(weight))

# Count by year, section, and trip
missing_counts <- missing_weight %>%
  count(year, section, trip)

# Plot with trip as factor
ggplot(missing_counts, aes(x = factor(trip), y = n, fill = section)) +
  geom_col(position = "dodge") +
  facet_wrap(~year) +
  labs(
    title = "Missing Weight Values by Year, Section, and Trip",
    x = "Trip (Factor)",
    y = "Count of Missing Values",
    fill = "Section"
  ) +
  theme_minimal()
```

## Rescaling the Data

If I wanted to rescale every quantitative variable in my dataset so that they
only have values between 0 and 1, I could use this formula:

</br>

$$y_{scaled} = \frac{y_i - min\{y_1, y_2,..., y_n\}}{max\{y_1, y_2,..., y_n\} 
- min\{y_1, y_2,..., y_n\}}$$

</br>

I might write the following `R` code to carry out the rescaling procedure for the `length` and `weight` columns of the `BlackfoorFish` data:

```{r}
#| echo: true
#| eval: false


fish <- trout_data |> 
  mutate(
    length = (length - min(length, na.rm = TRUE)) / 
             (max(length, na.rm = TRUE) - min(length, na.rm = TRUE)), 
    weight = (weight - min(weight, na.rm = TRUE)) / 
             (max(weight, na.rm = TRUE) - min(weight, na.rm = TRUE))
  )
```

This process of duplicating an action multiple times can make it difficult to
understand the intent of the process. *Additionally, it can make it very difficult to spot mistakes.*

**3. What is the mistake I made in the above rescaling code?**
min(length, na.rm = TRUE) in the denominator instead of min(weight, na.rm = TRUE)

When you find yourself copy-pasting lines of code, it's time to write a
function, instead!

**4. Transform the repeated process above into a `rescale_01()` function. Your function should...**

+ **... take a single vector as input.**
+ **... return the rescaled vector.**

```{r}
#| label: write-rescale-function
rescale_01 <- function(x) {
  # Input validation
  if (!is.numeric(x)) {
    stop("Input must be a numeric vector.")
  }
  if (length(x) <= 1) {
    stop("Input vector must have more than one element.")
  }

  # Efficient rescaling using range()
  r <- range(x, na.rm = TRUE)
  (x - r[1]) / (r[2] - r[1])
}
```

::: callout-tip
# Efficiency 

Think about the efficiency of the function you wrote. Are you calling the
**same** function multiple times? You might want to look into the `range()` 
function. 
:::

**5. Let's incorporate some input validation into your function. Modify your previous code so that the function stops if ...**

+ **... the input vector is not numeric.**
+ **... the length of the input vector is not greater than 1.**

::: callout-tip
# Modify Previous Code

Do not create a new code chunk here -- simply add these stops to your function
above!
:::

## Test Your Function

**6. Run the code below to test your function. Verify that the maximum of your rescaled vector is 1 and the minimum is 0!**

```{r}
#| label: verify-rescale-function

x <- c(1:25, NA)

rescaled <- rescale_01(x)
min(rescaled, na.rm = TRUE)
max(rescaled, na.rm = TRUE)
```

Next, let's test the function on the `length` column of the `BlackfootFish` data.

**7. The code below makes a histogram of the original values of `length`. Add a plot of the rescaled values of `length`. Output your plots side-by-side, so the reader can confirm the only aspect that has changed is the scale.**

::: callout-warning
This will require you to call your `rescale_01()` function within a `mutate()`
statement in order to create a `length_scaled` variable.
:::

```{r}
#| label: compare-original-with-rescaled-lengths
#| layout-ncol: 2

trout_scaled <- trout_data %>%
  mutate(length_scaled = rescale_01(length))

# Original length plot
ggplot(trout_scaled, aes(x = length)) +
  geom_histogram(binwidth = 45, fill = "steelblue") +
  labs(x = "Original Length (mm)", title = "Original Length") +
  scale_y_continuous(limits = c(0, 4000)) +
  theme_minimal()

# Rescaled length plot
ggplot(trout_scaled, aes(x = length_scaled)) +
  geom_histogram(binwidth = 0.05, fill = "darkorange") +
  labs(x = "Rescaled Length (0–1)", title = "Rescaled Length") +
  scale_y_continuous(limits = c(0, 4000)) +
  theme_minimal()

```

::: callout-tip
1. Set the y-axis limits for both plots to go from 0 to 4000 to allow for direct comparison across plots.

2. Pay attention to `binwidth`!

3. Use a Quarto code chunk option to put the plots side-by-side.
:::

## Challenge: Use Variables within a Dataset

Suppose you would like for your `rescale()` function to perform operations on a **variable within a dataset**. Ideally, your function would take in a data
frame and a variable name as inputs and return a data frame where the variable
has been rescaled.

**8. Create a `rescale_column()` function that accepts two arguments:**

+ **a dataframe**
+ **the name(s) of the variable(s) to be rescaled**

**The body of the function should call the original `rescale_01()` function you wrote previously. Your solution MUST use one of the `rlang` options from class.**

::: callout-tip
If you are struggling with this task, I recommend looking back over the 
[data frame functions](https://r4ds.hadley.nz/functions.html#data-frame-functions)
section of R for Data Science!
:::

```{r}
#| label: rescale-data-frame-function

```

**9. Use your `rescale_column()` function to rescale *both* the `length` and `weight` columns.**

::: callout-warning
I expect that you carry out this process by calling the `rescale_column()` function only ONE time!
:::

```{r}
#| label: rescale-two-columns

```
